---
title: "unused-text"
format: html
---


# From Data Structure


### Further Classifications of Attributes

While categorizing each attribute as categorical or quantitative is a good first step, statisticians and data analysts often go a step further and classify attributes based on their **scale of measurement**. This classification is based on how attributes were measured (i.e., how we assign numerals to objects of events) and what this implies about the empirical statements we can make about the constructs measured on that particular scale. The most common taxonomy for this was described by @Stevens:1946 who classified four scales of measurement: nominal, ordinal, interval, or ratio (NOIR). Below we describe each of these scales of measurement:

**Nominal Scale of Measurement:** In nominal scales of measurement, any numerals assigned to different values would only be useful as labels, for distinguishing or organizing values. Most categorical attributes have this scale of measurement. For example, in the NYT bests sellers data, the numerical values in the gender identity attribute (`female`) are only useful as labels and for distinguishing authors who identify as female and authors who don't. Because of this, only the following type of statement would be meaningful:

- In the NYT bestseller data, more authors identify as female (15) than do not identify as female (10).

The only type of empirical statements we can make are comparisons of the number of cases between different labels (e.g., counts, percentages).

**Ordinal Scale of Measurement:** Data measured using an ordinal scales of measurement, is still categorical. It has all the features of nominal measured data (e.g., labeling, distinguishing, organizing), but we can also rank order the values in a meaningful way. A classic example of the ordinal scale of measurement is in the 5-star review rating used on sites like Amazon. All of these statements would be meaningful:

- There are more 4-star reviews than 5-star reviews (comparison of counts).
- A review of 3 stars is better than a review of 2 stars (rank ordering).


With this scale of measurement, it is reasonable to not only provide counts of the different values (e.g., the number of 5-star reviews), but now because there is a rank ordering, we can also make empirical statements related to the rank ordering of the measured construct based on the numeral values. In nominal level data, these latter types of statements are not appropriate since the values for the labels are arbitrary. For example, authors with a gender identity of female were assigned a 1 and others were assigned a 0. This is arbitrary in that authors identifying as female could just as easily have been assigned a value of 0 and those that didn't identify as female a value of 1. This implies that even though the numeral 1 is greater than the numeral 0, the attribute values associated with these numerals (identifying as female or not) do not have any meaningful rank order. Because of this, saying something about one identity being greater than or less than another is inappropriate.


**Interval Scale of Measurement:** In interval level data, the rank order of the numbers assigned to attribute values is meaningful, similar to ordinal data. Moreover, the difference between consecutive values represents the same amount of difference on the underlying attribute. For example, consider the Fahrenheit temperature scale. All of these statements would be meaningful:

- There are more 30 degrees F days than 0 degrees F days (comparison of counts).
- A day that is 10 degrees F is warmer than a day that is 9 degrees F (rank ordering).
- The difference in temperature when you are comparing days that are 9 and 10 degrees F is the same as when you are comparing days that are 0 and 1 degrees F. (interval comparison).

The critical component is that on an interval scale, the difference in consecutive numerals has the same level of difference in the construct being measured, regardless of scale location. Again consider our 5-star rating system. The difference between a 1- and 2-star review is not the same as the difference between a 4- and 5-star review. While the numbers themselves have a constant difference, the difference in the amount of the underlying construct (satisfaction, happiness with the product, etc.) does not. With interval level scales we can compute summaries like means, standard deviations, and correlations.

**Ratio Scale of Measurement:** With attributes that have the ratio scale of measurement, the rank ordering of the numbers assigned to attribute values is meaningful, the differences between consecutive numerals indicates the same amount of difference in the underlying construct being measured, and ratio type statements about these differences are also meaningful. For example, the amount of snowfall is an attribute on the ratio scale of measurement. All of these statements would be meaningful:

- There are more days with 0 inches of snowfall than days with 10 inches of snowfall (comparison of counts).
- A day with 8 inches of snowfall got more snow than a day with 4 inches of snowfall (rank ordering).
- The difference in snowfall when you are comparing days with 8 inches of snowfall and 7 inches of snowfall is the same as when you are comparing days with 15 and 16 inches of snow (interval comparison).
- A day with 8 inches of snow got twice the amount of snow as a day with 4 inches of snow (ratio comparison).

Going back to our temperature scale, we cannot make these ratio type statements. For example, a day that is 60 degrees F is not twice as warm as a day that is 30 degrees F. This is because the Fahrenheit scale does not have a "true" zero value. (Zero degrees F does not indicate absence of temperature.) Whereas, in our snowfall attribute, a day with 0 inches of snow does indeed indicate no snow fell on that day.

Aside from the type of empirical statements we can make, the level of measurement also puts limits on the type of statistical analyses that are appropriate.

```{r}
#| label: tbl-measurement
#| tbl-cap: "The four measurement scales and the types of empirical statements and statistical summaries that are appropriate for each scale."
#| echo: false
library(gt)

tab_measurement = data.frame(
  scale = c("Nominal", "Ordinal", "Interval", "Ratio"),
  statement = c("Comparison of counts",
                "Comparison of counts <br /> Rank ordering",
                "Comparison of counts <br /> Rank ordering <br /> Interval comparisons",
                "Comparison of counts <br /> Rank ordering <br /> Interval comparisons <br /> Ratio comparisons"),
  stat = c("Counts, percentages",
           "Counts, percentages <br /> Median, percentiles",
           "Counts, percentages <br /> Median, percentiles <br /> Mean, standard deviation",
           "Counts, percentages <br /> Median, percentiles <br /> Mean, standard deviation <br /> Coefficient of variation")
)

tab_measurement |>
  gt() |>
  cols_label(
    scale = md("*Scale*"),
    statement = md("Empirical Statement"),
    stat = md("Statistical Summaries")
  ) |>
  cols_align(
    columns = c(scale, statement, stat),
    align = "left"
  ) |>
  fmt_markdown(columns = everything())

```

<br />


## How Were the Data Generated?

Another question that has direct implications on the methods used in data analysis is: How were the data generated or collected? Were they collected from a survey? Were they mined from the web? Were they generated as part of an experimental study? Knowing the answer to these questions also is important for the degree to which we can draw conclusions from the analysis.

Understanding how the data were generated allows us to determine whether the data we have constitute a **sample** of cases or the entire **population** of cases we are interested in learning about. Importantly, whether you have a sample or the entire population depends on how you define the cases/observations you are interested in drawing conclusions about.

:::fyi
**VOCABULARY**

A **population** includes *all* cases/observations of interest, whereas a **sample** includes a subset of cases from the population.
:::

For example, consider a child psychologist who wants to draw conclusions about all students at a particular school in Minnesota. To do this, she collects data from every student in that school. Since her data includes every case (student) she is interest in drawing conclusions for, her data would be a population, Now consider a second child psychologist who is interested in drawing conclusions about all students in Minnesota. This psychologist also collects data from every student in the same school as the first psychologist. This second psychologist's data would be considered a sample since the cases they included in their data are only a subset of the cases they want to draw conclusions about.

<br />



## Summary

Every time you encounter data, you should identify the cases and attributes in the data. Understanding the cases, especially in relation to the cases you want to draw conclusions about, helps you identify whether you have a sample, or the entire population. Classifying the attributes helps you think about the type of analysis you can undertake. If your data are a sample (rather than a population), you also need to ask how they were collected. Were they collected using randomness in the sampling method? Or is the sampling method used to collect the data biased?

<br />


## Experiments vs Observations

The type of analysis that is relevant for a given problem depends on the experimental design. More generally, experimental design refers to how the data were collected. How the data were collected has implications for the type of conclusions that can be drawn from the data (i.e., you may have heard the phrase, "correlation does not imply causation") and the subsequent analysis. A extremely sophisticated analysis does not overcome limitations in how the data were collected.

Given that how the data were collected greatly impacts the types of conclusions that can be drawn from the analysis, what are design features are important to consider? This topic is large, nuanced, and an entire series of courses have been created on this topic. The goal here is to think about some concepts that are particularly helpful to consider for any analysis.

The discussion will be framed around articulating whether the data collected were part of an experiment or were simply observed. The simplest definition for observational data are those that were collected without strong consideration about who is or how the data are collected. One example of observational data could be collected information about the shoes that people wear as they are walking in a busy part of town. Given this type of data collected, what could be some limitations about this type of data?

To contrast observational data, experimental data are those in which care was taken to how the data were collected. Within experimental data, it is common for there to be two or more conditions that are being explored. For example, in a clinical trial that is testing whether a vaccine or drug is effective and safe, the new vaccine or drug is often tested against a placebo. The placebo is a harmless substance that does not produce any change to the body, for example, the placebo could be a sugar pill that looks just like a new drug to be tested in every other way. The placebo would simply not contain the active ingredients of the new drug.

When conducting these experiments with two or more groups that are of interest to be compared, those who are participating in the study are randomly assigned or selected to be in one of the groups. Using the placebo vs new drug example, this would mean that each participant is randomly assigned to either receive the placebo or the new drug, but the participant does not know which one they are receiving. Often, those administering the treatment also do not know whether the placebo or active drug is being given as well.

<br />

### Explore Random Assignment

How does the random assignment of individuals to treatment conditions change the design from an observational to an experimental design? The random process is really the primary difference between being an observational vs an experiment. The random assignment to each condition in the study has the ability to, on average, even out differences across the two groups. Since the inclusion of being in one of the two groups is random, if the study has enough participants, it is more likely for the two groups to be as equal as possible across all characteristics for the study participants.

<!-- What could random assignment look like in practice? More to come ... .

<br />

### Example: Natural Experiments

<br />

## Data Structure

Data are often stored in tabular form for ease of use with common statistical programs, however data need not be in this structure. Data can come from anywhere and could consist of text, numbers representing some quantity, text labels representing groups, or many other formats. This section aims to give an introduction to the form and format of data, both common and uncommon.

<br />

### Tabular Data

Tabular data are those that are most commonly used in statistics courses. Tabular data are such that rows indicate unique cases of data and the columns represent different attributes for those cases. Table @tbl-tabular-data shows the an example of tabular data.

```{r}
#| label: tbl-tabular-data
#| tbl-cap: "Example of tabular data."
#| echo: false
#| message: false
#| warning: false
knitr::kable(
  data.frame(name = c('He-Man', 'She-Ra', 'Voltron'),
    human = c('Yes', 'Yes', 'No'),
    fictional = c('Yes', 'Yes', 'Yes'),
    height = c(83, 96, 3936)
    )
  )
```

In this table, each row represents a unique person and the first column would represent the identifying attribute for each unique row. For example, the first row represents characteristics for He-Man and each subsequent column represents specific attributes about them. For example, the second and third columns are Yes/No attributes indicating if the person is human or fictional or not. What are the primary differences between the second and third columns? Notice that the third column, the fictional attribute, is one in which all elements in the current tabular data are the same value. This would be an attribute that does not vary across the different rows in the data. These attributes are not helpful from a statistical perspective in this small data set, but if more data were added in which some elements were not fictional, then this attribute would contain useful information. In statistics terminology, this type of attribute would be called a **constant attribute**. In statistics, we are interested in attributes that **vary** across our units in the data. Attributes that vary are often referred to as **variables** in statistics terminology. Throughout this textbook, the term attribute will be used instead of variable. An attribute will refer to a column of data that carries information about the units in the data.

The final column of data in Table @tbl-tabular-data is one that represents the approximate height of each character, in inches ^[This is based on a search for the character name plus height, for example, "He-Man height."]. This attribute is different from the rest of the attributes in that it is a numeric quantity for each unit. Numeric quantities usually carry more information about magnitude differences compared to the text attributes described earlier. More explicitly, differences in the height attributes can be quantified and in this case, with the attribute being represented as the height in inches, each unit represents the same distance across the entire scale. That is, a one inch difference across the entire height scale is the same no matter where that inch occurs on the scale. These types of attributes will be used extensively in this text, most commonly these types of attributes will be used as the attribute of most interest in our analyses. For example, from the data above, it could be asked if there are differences in height of the characters based on if they are human or not. For this question, differences in height for those that are human (i.e., human attribute = "Yes") compared to those that are not human (ie., human attribute = "No") would be used as the primary comparison. This text will explore questions like this from a descriptive and inferential framework later.

:::exercises
For these data, there are only three rows, but you could imagine adding more rows to these elements for other characters. For example, imagine adding a new row to this data for a rabbit. Take a few minutes to think about what a new row with the rabbit would look like for each column of Table @tbl-tabular-data. Would any of the columns in the table change from a constant to now being a variable?
:::

<br />

### Non-tabular Data

Data can come in many different formats, this book will not extensively cover data that are not in a tabular format. However, non-tabular data are very common in practical problems. In these situations, these data are often wrangled into a more structured format to conduct a statistical analysis. Some common non-tabular data formats include data coming from text, video, audio, graphics, images, sensors, and even more.



